---
title: "Open Field and Y-maze"
author: "Bindoff, A., Sella, A."
date: "26 September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,results = "hold")

```

Behaviour data were supplied as an Excel file with a sheet for each experiment.

```{r libs}
library(ggplot2)  # for plotting
library(ggbeeswarm) # make beeswarm plots
library(reshape2) # to convert wide format data to long form (or vice versa)
library(dplyr)    # for transforming and summarising data more easily
library(openxlsx) # for reading in sheets from Excel files
library(lme4)     # for mixed effects models
library(lmerTest) # F-tests for mixed effects models fitted with lme4
```

## Open field

Load open field experiment data.  

```{r beam walk data}
of <- read.xlsx("nfl_open_field_and_y_maze_data.xlsx", sheet = 1)
of$genotype <- factor(of$genotype, levels = c("wt", "nfl_ko"))
of$month <- factor(of$month)
of$animal_id <- factor(of$animal_id)
head(of)
```

Filter out rows with a 1 in the `exclude` column.   


```{r exclusions1}
cat("\nrows before filter = ", nrow(of))  # check number of rows before we filter
of <- dplyr::filter(of, is.na(exclude))   # "filter all rows from bw where exclude = NA"
cat("\nrows after filter = ", nrow(of))   # check number of rows after we filter
```

The data were provided in long format so we don't need to "melt" it.  

We can plot the data to get a sense of what's happening, and also to check for any problems.  

```{r descriptive plots1}
ggplot(of, aes(x = month, y = distance_covered_in_.centre_cm, colour = genotype)) +
  #geom_boxplot() +
    geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Distance covered (cm)") +
  theme_bw()
```


```{r}
ggplot(of, aes(x = month, y = `velocity.cm/s`, colour = genotype)) +
  #geom_boxplot() +
    geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Velocity (cm/s)") +
  theme_bw()
```

```{r}
ggplot(of, aes(x = month, y = cumulative_duration_in_centre_from_centre_point_sec, colour = genotype)) +
  geom_boxplot() +
  #  geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Cumulative duration in centre (s)") +
  theme_bw()
```



### Analysis


```{r}
m1 <- lmer(distance_covered_in_.centre_cm ~ genotype + month + (1|animal_id), of)
round(summary(m1)$coefficients, 3)
```

Report 95% CIs and F-statistics  

```{r}
confint(m1)
anova(m1)
```

Check assumption of normality of residuals  

```{r}
qqnorm(resid(m1))
qqline(resid(m1))

```


```{r}
m2 <- lmer(`velocity.cm/s` ~ genotype + month + (1|animal_id), of)
round(summary(m2)$coefficients, 3)
confint(m2)
anova(m2)
```

```{r}
qqnorm(resid(m2))
qqline(resid(m2))
```

```{r}
m3 <- lmer(cumulative_duration_in_centre_from_centre_point_sec ~ genotype + month + (1|animal_id), of)
round(summary(m3)$coefficients, 3)
confint(m3)
anova(m3)
```

```{r}
qqnorm(resid(m3))
qqline(resid(m3))
```

## Y-maze

Load Y-maze experiment data.  

```{r y maze data}
ym <- read.xlsx("nfl_open_field_and_y_maze_data.xlsx", sheet = 2)
ym$genotype <- factor(ym$genotype, levels = c("wt", "nfl_ko"))
ym$month <- factor(ym$month)
ym$animal_id <- factor(ym$animal_id)
head(ym)
```

Filter out rows with a 1 in the `exclude` column.   


```{r exclusions2}
cat("\nrows before filter = ", nrow(ym))  # check number of rows before we filter
ym <- dplyr::filter(ym, is.na(exclude))   # "filter all rows from bw where exclude = NA"
cat("\nrows after filter = ", nrow(ym))   # check number of rows after we filter
```

The data were provided in long format so we don't need to "melt" it.  

We can plot the data to get a sense of what's happening, and also to check for any problems.  


```{r}
ggplot(ym, aes(x = month, y = velocity_cm_s, colour = genotype)) +
  #geom_boxplot() +
    geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Velocity (cm/s)") +
  theme_bw()
```

Looks like a couple of outliers, here's how to remove them (if you should?)  

```{r}
ym <- dplyr::filter(ym, velocity_cm_s < 50)
```
```{r}
ggplot(ym, aes(x = month, y = velocity_cm_s, colour = genotype)) +
  #geom_boxplot() +
    geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Velocity (cm/s)") +
  theme_bw()
```


```{r descriptive plots2}
ggplot(ym, aes(x = month, y = distance_travelled_cm, colour = genotype)) +
  #geom_boxplot() +
    geom_quasirandom() +
  scale_colour_manual(values = c("black", "red")) +
  ylab("Distance travelled (cm)") +
  theme_bw()
```

Outlier?  



### Analysis


```{r}
m1 <- lmer(velocity_cm_s ~ genotype + month + (1|animal_id), ym)
round(summary(m1)$coefficients, 3)
confint(m1)
anova(m1)
```


```{r}
qqnorm(resid(m1))
qqline(resid(m1))

```


```{r}
m2 <- lmer(distance_travelled_cm ~ genotype + month + (1|animal_id), ym)
round(summary(m2)$coefficients, 3)
confint(m2)
anova(m2)
```

```{r}
qqnorm(resid(m2))
qqline(resid(m2))
```

```{r}
m3 <- lmer(cumulative_duration_novel_arm_sec  ~ genotype + month + (1|animal_id), ym)
round(summary(m3)$coefficients, 3)
confint(m3)
anova(m3)
```

```{r}
qqnorm(resid(m3))
qqline(resid(m3))
```


And so on...   

